---
title: "TSAR Project Assignment Part1 - Group 3"
format: html
editor: visual
---

## Introduction

## Preliminary tasks

This analysis begins by establishing the necessary computational environment in R and importing the required dataset. First, the script ensures that essential data manipulation libraries, specifically `data.table` and `dplyr`, are installed and loaded. Following this setup, the code loads the working dataset named `myLCdata.csv`. The commented code outlines the reproducible generation of this working file from the original `LCdata.csv` dataset. By setting a fixed random seed, the script extracted a 50 percent random sample of the observations and selected 5 of the 19 available features.

### Package installation

```{r}
# install the required packages if needed
if (!require("data.table")) install.packages("data.table")
if (!require("dplyr")) install.packages("dplyr")
# load them
library(data.table)
library(dplyr)
```

### Loading data

```{r}
# Random sample your group's data set from the original LCdata data set
#LCdata <- fread("LCdata.csv")
#set.seed(3)
#myLCdata <- LCdata %>% sample(5) %>% slice_sample(prop = .5)
#fwrite(myLCdata, file = "myLCdata.csv")
myLCdata <- fread("myLCdata.csv")
```

## Tasks

### Task 1. Business Understanding

```{r}
glimpse(myLCdata)
str(myLCdata)
# Extracting the first 6 rows of the data to get an understanding
head(myLCdata)
```

**Data mapping**

+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Name**             | **Description**                                                                                                                                                                                                              |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`annual_inc`**     | The self-reported annual income provided by the borrower during registration.                                                                                                                                                |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`home_ownership`** | The home ownership status provided by the borrower during registration. Our values are: RENT, OWN, MORTGAGE, OTHER.                                                                                                          |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`inq_last_6mths`** | The number of inquiries in past 6 months (excluding auto and mortgage inquiries)                                                                                                                                             |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`open_acc`**       | The number of open credit lines in the borrower's credit file.\                                                                                                                                                              |
|                      | Remark:\                                                                                                                                                                                                                     |
|                      | - A credit line is a type of loan that allows an individual or business to borrow money and repay it, often on a revolving basis without applying for a new loan.\                                                           |
|                      | - A revolving credit is a credit line that remains available even as you pay the balance. Borrowers can access credit up to a certain amount and then have ongoing access to that amount of credit. Typically a credit card. |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`purpose`**        | A category provided by the borrower for the loan request.                                                                                                                                                                    |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

### Task 2. Data Understanding Based on Summary Statistics

### Task 3. Univariate Exploratory Data Analysis (EDA) with ggplot2

### Task 4. Bivariate EDA with ggplot2
