---
title: "TSAR Project Assignment Part1 - Group 3"
format: html
editor: visual
execute: 
  warning: false
---

## Introduction

## Preliminary tasks

This analysis begins by establishing the necessary computational environment in R and importing the required dataset. First, the script ensures that essential data manipulation libraries, specifically `data.table` and `dplyr`, are installed and loaded. Following this setup, the code loads the working dataset named `myLCdata.csv`. The commented code outlines the reproducible generation of this working file from the original `LCdata.csv` dataset. By setting a fixed random seed, the script extracted a 50 percent random sample of the observations and selected 5 of the 19 available features.

### Package installation

```{r}
# install the required packages if needed
if (!require("data.table")) install.packages("data.table")
if (!require("dplyr")) install.packages("dplyr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("GGally")) install.packages("GGally")


# load them
library(data.table)
library(dplyr)
library(ggplot2)
library(GGally)

# Show the scales as human readable numbers rather than scientific ones
options(scipen = 999)
```

### Loading data

```{r}
# Random sample your group's data set from the original LCdata data set
#LCdata <- fread("LCdata.csv")
#set.seed(3)
#myLCdata <- LCdata %>% sample(5) %>% slice_sample(prop = .5)
#fwrite(myLCdata, file = "myLCdata.csv")
myLCdata <- fread("myLCdata.csv")
```

## Tasks

### Task 1. Business Understanding

```{r}
glimpse(myLCdata)
str(myLCdata)
# Extracting the first 6 rows of the data to get an understanding
head(myLCdata)
```

**Data mapping**

+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Name**             | **Description**                                                                                                                                                                                                              |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`annual_inc`**     | The self-reported annual income provided by the borrower during registration.                                                                                                                                                |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`home_ownership`** | The home ownership status provided by the borrower during registration. Our values are: RENT, OWN, MORTGAGE, OTHER.                                                                                                          |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`inq_last_6mths`** | The number of inquiries in past 6 months (excluding auto and mortgage inquiries)                                                                                                                                             |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`open_acc`**       | The number of open credit lines in the borrower's credit file.\                                                                                                                                                              |
|                      | Remark:\                                                                                                                                                                                                                     |
|                      | - A credit line is a type of loan that allows an individual or business to borrow money and repay it, often on a revolving basis without applying for a new loan.\                                                           |
|                      | - A revolving credit is a credit line that remains available even as you pay the balance. Borrowers can access credit up to a certain amount and then have ongoing access to that amount of credit. Typically a credit card. |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`purpose`**        | A category provided by the borrower for the loan request.                                                                                                                                                                    |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

### Task 2. Data Understanding Based on Summary Statistics

```{r}
summary(myLCdata)
skimr::skim(myLCdata)
```

### Task 3. Univariate Exploratory Data Analysis (EDA) with ggplot2

Now we want to shift from Data Understanding to Data Exploration. Because we have numerical and categorical data, we'll use plots respectively.

3.1 Numerical Variables

-   annual_inc

-   open_acc

-   inq_last_6mths

3.2 Categorical Variables

-   home_ownership

-   purpose

```{r}


# 3.1 annual_inc
ggplot(myLCdata, aes(x = annual_inc)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") + 
  scale_x_log10() + 
  labs(title = "Distribution of Annual Income (Log Scale)", x = "Annual Income", y = "Count")

ggplot(myLCdata, aes(y=annual_inc)) +
  geom_boxplot() + 
  scale_y_log10() + 
  labs(y="Annual income (Log scale)", title="Boxplot visualisation of the annual_inc variable")

# Q1
quantile(myLCdata$annual_inc, .25)

# median
median(myLCdata$annual_inc)

# Q3
quantile(myLCdata$annual_inc, .75)
```

The upper visualizations show the distribution of the annual income based on a logarithmic scale. Using the additional functions `quantile()` and `median()` we can explore how the data is distributed. As seen the first quartile (25%) is 45649, the median is 65000 and the third quartile (75%) is 90000. Having these numbers it's visible, that 75% of loan applicants earn an annual income of 90000 and less, whereas only 25% earn an income greater than 90000. Numerous outliers are visible both above and below the whiskers. The scale reaches over 1,000,000 in the upper region. These extreme values are typical for income data because the distribution of salaries and incomes vary depending on gender, occupation and work experience

```{r}
# 3.1 open_acc
ggplot(myLCdata, aes(x = open_acc)) + 
  geom_histogram(binwidth = 1, fill = "darkorange", color = "black", alpha = 0.8) + 
  labs(title = "Distribution of Open Credit Lines", 
       x = "Number of Open Accounts",
       y = "Frequency") +
  theme_minimal()
```

```{r}
# 3.1 inq_last_6mths
ggplot(myLCdata, aes(x=inq_last_6mths)) +
  geom_bar(fill="seagreen", color="black", alpha=0.8) +
  labs(title = "Inquiries in the Past 6 Months",
       x = "Number of Inquiries", 
       y = "Frequency") + 
  theme_minimal()

ggplot(myLCdata, aes(y=inq_last_6mths)) + 
  geom_boxplot() + 
  scale_y_log10()

# Q1
quantile(myLCdata$inq_last_6mths, .25, na.rm=TRUE)

# median
median(myLCdata$inq_last_6mths, na.rm=TRUE)

# Q3
quantile(myLCdata$inq_last_6mths, .75, na.rm=TRUE)

myLCdata %>% group_by(inq_last_6mths) %>% summarise(count = n()) %>% arrange(-count)
```

The variable inq_last_6mths shows the number of times one and the same person applied or inquired for a loan application. The lower the number, best case scenario would be 0, shows, that this person hasn't applied for another loan in the past 6 months, meaning that this person could theoetically be financially stable. A higher number could mean, that this person might be considered a risky customer, as they perhaps cannot repay the other loans, and therefore might be denied the loan application.

The trend in the data is that the loan applicants have either applied for 0 loans, 1 loan or 2 loans in the 6 month period. This shows, that customers can apply while having an ongoing loan application. The outliers in this column are the customers, that have applied for multiple loans in the past 6 months, in our case the highest number of inquiries in the past 6 months are 10, 27 and 31, having each one customer. These three customers have therefore inquired for multiple loans in the past 6 months, and could therefore be classified as higher risk customers.

```{r}
# 3.2 purpose
ggplot(myLCdata, aes(x = purpose)) + 
  geom_bar(fill = "tomato") + 
  coord_flip() + 
  labs(title = "Frequency of Loan Purposes")

myLCdata %>% group_by(purpose) %>% summarise(count=n()) %>% arrange(-count)
```

The diagram shows the overall distribution of the loan purposes. Using a `group_by` pipeline, it can be seen that debt consolidation is the primary reason why the loan has been requested.

To understand what debt consolidation is, we must take a look at the definition: \
"Debt consolidation is a form of debt refinancing that entails taking out one loan to pay off many others" (Wikipedia: Debt consolidation, 2026).

This form of loan is common, as short term loans can have high interest rates, therefore, applicants may request a loan with a longer term, where the interest is significantly lower compared to short term loans.

### Task 4. Bivariate EDA with ggplot2

Now we are looking for some obvious correlations or patterns of variables.

**4.1 First impression with ggpairs()**

```{r}
GGally::ggpairs(myLCdata %>% select(annual_inc, open_acc, purpose, home_ownership, inq_last_6mths))
```

In this ggpairs matrix we can see that the numeric variables (annual_inc, open_acc and inq_last_6mths) are heavily right-skewed. Furthermore these variables show extreme outliers, clustering most data points near zero.

Additionally we see that the linear correlations between these numeric variables are very weak. The "highest" correlation we can see is 0.151 between annual income and open accounts.

This suggests that we need to transform the data using a log scale.

**4.2 Numeric vs. Categorical**

-   home_ownership vs. annual_inc

```{r}
ggplot(myLCdata, aes(x = home_ownership, y = annual_inc, fill = home_ownership)) + 
  geom_boxplot(alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Annual Income by Home Ownership Status",
       x = "Home Ownership",
       y = "Annual Income (Log Scale)") + 
  theme_minimal() +
  theme(legend.position = "none")
```

In this boxplot we see a clear hierarchy in median annual income based on housing status. Borrowers with a 'Other' or 'Any' category demonstrate the highest median incomes. These are followed by those who have a 'Mortage' on their homes and the ones from the category 'Own'. The 'Rent' and 'None' categories have the lowest median incomes.

The log scale emphasizes that the 'Mortage' and 'Rent' categories contain the highest concentration of extreme upper-income outliers. We see this because of how high the black dots stretch upwards. The points at the very top represent extremely high incomes in the millions.

To get a clear ranking of the 'Home Ownership' categories, we created the following ranking.

```{r}
myLCdata %>% group_by(home_ownership) %>%
  summarise(Median_Income = median(annual_inc, na.rm = TRUE)) %>%
  arrange(desc(Median_Income))
```

**4.3 Numeric vs. Numeric**

-   open_acc vs. inq_last_6mths

We use geom_jitter to add tiny amounts of random noise to the placement of the dots. We also add a trend line to see if there is any relationship.

```{r}
ggplot(myLCdata, aes(x = open_acc, y = inq_last_6mths)) +
  geom_jitter(alpha = 0.3, color = "purple", width = 0.25, height = 0.25) +
  geom_smooth(method = "lm", color = "black", se = FALSE) + 
  labs(title = "Open Accounts vs. Recent Inquiries",
       subtitle = "Points are jittered to reveal data density",
       x = "Number of open Accounts",
       y = "Inquiries in the Past 6 Months") + 
  theme_minimal()
```

The plotted trend line indicates a slight positive relationship between a borrower's number of open credit accounts and the number of credit inquiries they have made in past 6 months. This matches our insight from our ggpairs matrix in 4.1 where we saw that these two attributes have a weak correlation of 0.124.

Now this slight positive correlation means that individuals with a higher number of credit lines might be slightly more active in seeking new credit. But most borrowers have very few recent inquiries regardless of their total open accounts.

**4.4 Categorical vs. Categorical**

-   purpose vs. home_ownership

We want to see if loan purposes vary based on living situations. We use a 100% Stacked Bar Chart.

```{r}
ggplot(myLCdata, aes(x = purpose, fill = home_ownership)) + 
  # position = fill standardizes the bars to 1.0
  geom_bar(position = "fill", color = "white", linewidth = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) + 
  labs(title = "Proportion of Home Ownership by Loan Purpose",
       x = "Loan Purpose",
       y = "Percentage",
       fill = "Home Ownership") +
  theme_minimal() + theme(legend.position = "bottom")

```

Lets look at this graph ordered by home_ownership == 'RENT', to better compare the different categories.

```{r}
ggplot(myLCdata, aes(x = forcats::fct_reorder(purpose, home_ownership == "RENT", mean), fill = home_ownership)) + 
  # position = fill standardizes the bars to 1.0
  geom_bar(position = "fill", color = "white", linewidth = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) + 
  labs(title = "Proportion of Home Ownership by Loan Purpose",
       x = "Loan Purpose",
       y = "Percentage",
       fill = "Home Ownership") +
  theme_minimal() + theme(legend.position = "bottom")

```

By ordering the stacked bars by the proportion of the renters, a behavioral divide becomes visible.

At the top of the chart life events such as 'moving', 'educational' and 'wedding' are overwhelmingly driven by renters.

In the middle standard financial product such as 'car', 'credit card' and 'small business' show a balanced near 50/50 split between renters and mortage holders.

At the bottom we see the reversal from above, where 'home_improvement' loans are almost entirely taken out by mortage holders. This makes sense in our opinion, because their incentive to care for their immobilia is higher than the one for renters.

**4.5 Adding a 3rd Variable**

We will use facet_wrap

```{r}
ggplot(myLCdata) + geom_point(aes(x=annual_inc, y=inq_last_6mths)) + scale_x_log10() + scale_y_log10() + facet_wrap(~home_ownership)

ggplot(myLCdata) + geom_point(aes(x=annual_inc, y=inq_last_6mths)) + scale_x_log10() + scale_y_log10() + facet_wrap(~purpose)
```
