---
title: "TSAR Project Assignment Part1 - Group 3"
format: html
editor: visual
---

## Introduction

## Preliminary tasks

This analysis begins by establishing the necessary computational environment in R and importing the required dataset. First, the script ensures that essential data manipulation libraries, specifically `data.table` and `dplyr`, are installed and loaded. Following this setup, the code loads the working dataset named `myLCdata.csv`. The commented code outlines the reproducible generation of this working file from the original `LCdata.csv` dataset. By setting a fixed random seed, the script extracted a 50 percent random sample of the observations and selected 5 of the 19 available features.

### Package installation

```{r}
# install the required packages if needed
if (!require("data.table")) install.packages("data.table")
if (!require("dplyr")) install.packages("dplyr")
# load them
library(data.table)
library(dplyr)
```

### Loading data

```{r}
# Random sample your group's data set from the original LCdata data set
#LCdata <- fread("LCdata.csv")
#set.seed(3)
#myLCdata <- LCdata %>% sample(5) %>% slice_sample(prop = .5)
#fwrite(myLCdata, file = "myLCdata.csv")
myLCdata <- fread("myLCdata.csv")
```

## Tasks

### Task 1. Business Understanding

```{r}
glimpse(myLCdata)
str(myLCdata)
# Extracting the first 6 rows of the data to get an understanding
head(myLCdata)
```

**Data mapping**

+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **Name**             | **Description**                                                                                                                                                                                                              |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`annual_inc`**     | The self-reported annual income provided by the borrower during registration.                                                                                                                                                |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`home_ownership`** | The home ownership status provided by the borrower during registration. Our values are: RENT, OWN, MORTGAGE, OTHER.                                                                                                          |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`inq_last_6mths`** | The number of inquiries in past 6 months (excluding auto and mortgage inquiries)                                                                                                                                             |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`open_acc`**       | The number of open credit lines in the borrower's credit file.\                                                                                                                                                              |
|                      | Remark:\                                                                                                                                                                                                                     |
|                      | - A credit line is a type of loan that allows an individual or business to borrow money and repay it, often on a revolving basis without applying for a new loan.\                                                           |
|                      | - A revolving credit is a credit line that remains available even as you pay the balance. Borrowers can access credit up to a certain amount and then have ongoing access to that amount of credit. Typically a credit card. |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+
| **`purpose`**        | A category provided by the borrower for the loan request.                                                                                                                                                                    |
+----------------------+------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------+

### Task 2. Data Understanding Based on Summary Statistics

```{r}
summary(myLCdata)
skimr::skim(myLCdata)
```

### Task 3. Univariate Exploratory Data Analysis (EDA) with ggplot2

Now we want to shift from Data Understanding to Data Exploration. Because we have numerical and categorical data, we'll use plots respectively.

3.1 Numerical Variables

-   annual_inc

-   open_acc

-   inq_last_6mths

3.2 Categorical Variables

-   home_ownership

-   purpose

```{r}
library(ggplot2)

# 3.1 annual_inc
ggplot(myLCdata, aes(x = annual_inc)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") + 
  scale_x_log10() + 
  labs(title = "Distribution of Annual Income (Log Scale)", x = "Annual Income", y = "Count")

# 3.1 open_acc
ggplot(myLCdata, aes(x = open_acc)) + 
  geom_histogram(binwidth = 1, fill = "darkorange", color = "black", alpha = 0.8) + 
  labs(title = "Distribution of Open Credit Lines", 
       x = "Number of Open Accounts",
       y = "Frequency") +
  theme_minimal()

# 3.1 inq_last_6mths
ggplot(myLCdata, aes(x=inq_last_6mths)) +
  geom_bar(fill="seagreen", color="black", alpha=0.8) +
  labs(title = "Inquiries in the Past 6 Months",
       x = "Number of Inquiries", 
       y = "Frequency") + 
  theme_minimal()

# 3.2 purpose
ggplot(myLCdata, aes(x = purpose)) + 
  geom_bar(fill = "tomato") + 
  coord_flip() + 
  labs(title = "Frequency of Loan Purposes")
```

### Task 4. Bivariate EDA with ggplot2

Now we will look at how the variables interact. Here we should find the story in our data.

4.1 Numeric vs. Categorical

-   home_ownership vs. annual_inc

```{r}
ggplot(myLCdata, aes(x = home_ownership, y = annual_inc, fill = home_ownership)) + 
  geom_boxplot(alpha = 0.7) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Annual Income by Home Ownership Status",
       x = "Home Ownership",
       y = "Annual Income (Log Scale)") + 
  theme_minimal() +
  theme(legend.position = "none")
```

4.2 Numeric vs. Numeric

-   open_acc vs. inq_last_6mths

We use geom_jitter to add tiny amounts of random noise to the placement of the dots. We also add a trend line to see if there is any relationship.

```{r}
ggplot(myLCdata, aes(x = open_acc, y = inq_last_6mths)) +
  geom_jitter(alpha = 0.3, color = "purple", width = 0.25, height = 0.25) +
  geom_smooth(method = "lm", color = "black", se = FALSE) + 
  labs(title = "Open Accounts vs. Recent Inquiries",
       subtitle = "Points are jittered to reveal data density",
       x = "Number of open Accounts",
       y = "Inquiries in the Past 6 Months") + 
  theme_minimal()
```

4.3 Categorical vs. Categorical

-   purpose vs. home_ownership

We want to see if loan purposes vary based on living situations. We use a 100% Stacked Bar Chart.

```{r}
ggplot(myLCdata, aes(x = purpose, fill = home_ownership)) + 
  # position = fill standardizes the bars to 1.0
  geom_bar(position = "fill", color = "white", linewidth = 0.2) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) + 
  labs(title = "Proportion of Home Ownership by Loan Purpose",
       x = "Loan Purpose",
       y = "Percentage",
       fill = "Home Ownership") +
  theme_minimal() + theme(legend.position = "bottom")

```
